Using an SMT engine to generate Symbolic Automata \thanks{This work was partially funded by the Associated Team FM4CPS between INRIA and ECNU, Shanghai}

### 2019-03-21 2017-Xudong-SMT/ActaInformatica.tex
Correct the format of IfThenElse.
Modify the \algnewcommand{\IfThenElse}.
Use \quad as the indent for the ITE statements.
Polish Enable example, replace figures accordingly

\newcommand{\sm}[1]{\mbox{\boldmath\small #1}}
\usepackage{algorithm} 
%\usepackage{algorithmicx} 
\usepackage{algpseudocode}  
\renewcommand{\algorithmicrequire}{\textbf{Input:}} 
\renewcommand{\algorithmicensure}{\textbf{Output:}}
\algnewcommand{\IfThen}[2]{\State \algorithmicif\ #1\ \algorithmicthen\ #2}% \IfThenElse{<if>}{<then>}
\algnewcommand{\IfThenElse}[3]{\State \algorithmicif\ #1\ \algorithmicthen\ #2\ \algorithmicelse\ #3}% \IfThenElse{<if>}{<then>}{<else>}
%\DeclareMathOperator{\card}{card}
%\DeclareMathOperator{\Flat}{Flat}
\usepackage{listings}

@@

\paragraph{Structure.}
In section
\ref{section:pnets} we give a description and a formal definition of
the pNet model, as found in previous publications. Then in section
\ref{section:examples} we present our use-case, based on a BIP
architecture from the nano-satellite case-study. 
Section \ref{section:op-semantics} recalls briefly the operational semantics
of pNet.
Section \ref{section:implementation} explains in details the algorithm
used to compute this semantics, including the interaction with Z3, and
shows the full result of the semantic computation on the running example.
Finally we conclude and discuss perspectives in Section
\ref{section:conclusion}.

@@

  \includegraphics[width=\linewidth]{XFIG/Enable1}
  \\[1.3ex]
 \includegraphics[width=\linewidth]{XFIG/Enable2}
  \caption{Two pNet encodings for  Enable }  \label{schema:enable-pnets}
\end{minipage}
  \hspace{2mm}
\begin{minipage}{6cm}
  \includegraphics[width=\linewidth]{XFIG/P-QR}
  \caption{Composed pNet for ``P>>(Q>>R)''}  \label{schema:enable-composed}
\end{minipage}

@@

	To give simple intuitions of the open pNet model and its semantics, we use
	here a simple example based on the ``Enable'' operator of the LOTOS language.
	In Fig. \ref{schema:enable-pnets}, we show 2 possible encodings of the  
	``Enable'' operator.
	In the Enable expression ``P>>Q'', an \texttt{exit(x)} statement within P
	terminates the current process, carrying a value $x$ that
	is captured by the \texttt{accept(x)} statement of Q.

	We use a simple action algebra,
	containing two constructors $\delta(x)$ and $acc(x)$, for any
	possible data type of the variable $x$, corresponding to the
	statements \texttt{exit(x)} and \texttt{accept(x)}. Both $\delta(x)$
	and $acc(x)$ actions are
	implicitly included in the sorts of all processes.
	We need no specific predicate over the action expressions, apart
	from equality of actions. 
	In the first encoding $Enable1$, in the upper part of
	Fig. \ref{schema:enable-pnets}, we use a controller $C_1$ with two
	states, and simple control actions $l, r, \delta$. The second encoding
	$Enable2$ uses a data-oriented style, with a single state controller, and a
	state-variable $s_0$, with values in $\{0,1\}$.

@@

	In Fig. \ref{schema:enable-composed}, we enrich our example by
	composing 2 occurences of the Enable1 pNet. To simplify,
	we only have represented one instance of the synchronisation vector set,
	and of the controller.

@@

\begin{example}[Open transitions]
  Transition $OT_1$ of the pNet $EnableCompl$ is generated by
  combination of the first vector of its root node, with
  the second vector of its $2^{nd}$ node. Its generated predicate is
  trivialy true. The global states shows that controler $C_3$ has
    changed from state 0 to 1.

  \label{OT:example-1}
        \begin{mathpar}
	  OT_1  =
          \inferrule*[myfraction=\reddottedrule]
	{\{\xrightarrow{\delta(x3)}_P ~~
          \xrightarrow{acc(x3)}_Q \} ~~
          [\forall z. acc(x3) \neq \delta(z)]
        }
	{\ostate{0,0} \xrightarrow {\underline{\delta(x3)}} \ostate{1,0}}
	\end{mathpar}

If we use the data-oriented pNet Enable2 from Figure \ref{schema:enable-pnets} to building a similar
``P>>(Q>>R)'' expression, the corresponding open transition would be:

  \label{OT:example-1}
        \begin{mathpar}
	  OT_2  =
          \inferrule*[myfraction=\reddottedrule]
	{\{\xrightarrow{\delta(x3)}_P ~~
          \xrightarrow{acc(x3)}_Q \} ~~
          [v_1=0 \land v_2 = 0 \land \forall z. acc(x3) \neq
            \delta(z)] ~~ \{v1:=1\}
        }
	{\ostate{0,0} \xrightarrow {\underline{\delta(x3)}} \ostate{0,0}}
	\end{mathpar}
        where $v_1$ and $v_2$ are respectively the state variables of
        the first and second Enable2 pNet. Indeed the global
        state space of this version is reduced to $\ostate{0,0}$, and
        the state change is replaced by the assignment ``$v_1:=1$''.

@@

\begin{figure}[t]
\begin{mathpar}
    ot = \openrule{\{s0 \xrightarrow{fail(true)} s0, t0 \xrightarrow{resume(false)} t0\}, ~~\{\xrightarrow{hb}\},~~fail(true)=fail(b_1)\\
    \OTland resume(false)=start(b_2)\OTland hb=fail(b_0)\OTland v=\underline{fail}\OTland b_1=b_2\\
    \OTland (b_1\lor b_2) \Rightarrow b_0, ~~\{\}}
    	   {\ostate{s0,t0} \xrightarrow{v} \ostate{s0,t0}}
  \end{mathpar}
\caption{One of the unsatisfiable open transitions \TODO{to be changed}}  \label{schema:unsat-ot}
\end{figure}

Our matching/filtering strategy builds some transitions where
the predicates express incompatible constraints. Even if having an
unsatisfiable (symbolic) transition would not be incorrect,
we choose to minimize the open automaton (\ie its number of
transitions and states), by checking the predicates for satisfiability.
In Fig.~\ref{schema:unsat-ot}, we show an 
unsatisfiable open transition 
from the open automaton of our running example. It shows the case where
the failure controller performs a ``$\PortFail$'' action, while the timer
executes a ``$\PortResume$''. The chosen synchronization
vector (SV0 from Fig.~\ref{schema:ArchFailure:pNet}) does not match
with these actions, since it expects $\NameTimer.\PortStart$. This
  mismatch is materialised by the predicate
fragment ``$\mathit{resume}(\false)=\mathit{start}(b_2)$''.

@@

\begin{algorithm}[h]
  \caption{``Smart'' Reachable Open Automaton Generation}
  \label{alg3}
\begin{algorithmic}[1]
\Require A pNet $P$ (cannot be a hole)
        %\qquad \hl{\textbackslash\textbackslash~SB: was `` pn ''}
\State Initialize sets $U=\{\mathit{InitState}(P)\}$ and $E=\emptyset$,
for unexplored and explored global states, respectively; $L=\emptyset$ for the resulting OTs;
\While{!$\mathit{isEmpty}(U)$}
	\State Choose $S$ in $U$; remove $S$ from $U$, add $S$ to $E$;
        %\qquad \hl{\textbackslash\textbackslash~SB: was `` Choose $S$ in $U$ ''}
	\State $\mathit{OTs} = \mathit{MakeTransitions}(P, S)$;
%%        \State Store $\mathit{OT}$ in $L$; 
	\For {each $\mathit{OT} \in \mathit{OTs}$}
        Check satisfiability of $\mathit{OT}$ using the SMT solver;
        \IfThen {$SAT(\mathit{OT})$}
                {\State \{Add $\mathit{OT}$ to $L$;
                  \State Let $S'$ be the target global state of $\mathit{OT}$
	          \IfThen{($S' \not\in U \cup E$)}{Add $S'$ into $U$;}
                  \}}
	\EndFor
\EndWhile
%%\State Prune the unsatisfiable open transitions in $L$ using the SMT solver;
\IfThenElse {(NonEmptyAssignmentSet($\mathit{L}$))}
        {\State $\mathit{L1}$=filterUnsatTransitions($\mathit{L,Assignments}$);
          \State \Return $\mathit{OA}$=makeAutomaton($\mathit{InitState}(P),L)$;}
        {\State \Return $\mathit{OA}$=makeReachableSubAutomaton($\mathit{InitState}(P),L1)$};
\end{algorithmic}  
\end{algorithm}

@@

Figure~\ref{schema:smt-lib} shows the translation of the
transition of Fig.~\ref{schema:unsat-ot}
in the SMTlib syntax.
It contains the declaration
of the BIP action algebra sorts 
and constructors, then the declaration of variables, and finally the
predicate to be checked, encoded as a set of assertions.
Here the behavior of the hole ``B'' appears as a variable
``|hb\_B:13:1|'', and line 10 shows the constraint that matching with
a synchronisation vector has imposed on the behavior of ``B''.
Here we display also the diagnosis (``sat'' or ``unsat'') generated by Z3.

\begin{figure}[t]
  %  \centerline{\includegraphics[width=12cm]{XFIG/SMTLIB2}}
    \centerline{\includegraphics[width=\linewidth]{XFIG/SMTLIB3}}
  \caption{The input of the Z3 solver in SMT-LIB language and the output result\TODO{Change}}  \label{schema:smt-lib}
\end{figure}

@@

\section{Use-cases}
\TODO{new section here, showing various features on different examples}

\subsection{Enable operator} in Lotos, with (equivalent) state-based and data-based encodings; Associativity of Enable;

\subsection{simple communication protocol} specification and implementation;

\subsection{BIP FailureTimer architecture}

\label{section:FailureTimerMax}
\label{section:BIParchitectures}
\label{section:examples}

@@

\subsection{Composing 2 FailureTimer BIP Architectures}
\noteInEM{Modifications stopped here}
\label{section:resultTimerInterval}
\TODO{Enhanced FailureTimer, 2 instances composed, with duplicated control and Timer}

>>>>

\newcommand{\sm}[1]{\mbox{\boldmath\small #1}}
\usepackage{algorithm} 
%\usepackage{algorithmicx} 
\usepackage{algpseudocode}  
\renewcommand{\algorithmicrequire}{\textbf{Input:}} 
\renewcommand{\algorithmicensure}{\textbf{Output:}}
\algnewcommand{\IfThen}[2]{\State \algorithmicif\ #1\ \algorithmicthen\ #2}% \IfThenElse{<if>}{<then>}
\algnewcommand{\IfThenElse}[3]{\State \algorithmicif\ #1\ \algorithmicthen\ #2\ \State \algorithmicelse\ #3}% \IfThenElse{<if>}{<then>}{<else>}
%\DeclareMathOperator{\card}{card}
%\DeclareMathOperator{\Flat}{Flat}
\usepackage{listings}

@@

\paragraph{Structure.}
In section
\ref{section:pnets} we give a description and a formal definition of
the pNet model, as found in previous publications, and present a
running example based on Lotos ``Enable'' operator.
Section \ref{section:op-semantics} recalls briefly the operational semantics
of pNet.
Section \ref{section:implementation} explains in details the algorithm
used to compute this semantics, including the interaction with Z3, and
shows the full result of the semantic computation on the running
example.
In section \ref{section:use-cases} we show the result of the study of
our running example, then explain a bigger use-case based on a BIP
architecture from the nano-satellite case-study. We show the benefits
drawn from the smart algorithm.
Finally we conclude and discuss perspectives in Section
\ref{section:conclusion}.

@@

  \includegraphics[width=\linewidth]{ActaXFIG/Enable1}
  \\[1.3ex]
 \includegraphics[width=\linewidth]{ActaXFIG/Enable2}
  \caption{Two pNet encodings for  Enable }  \label{schema:enable-pnets}
\end{minipage}
  \hspace{2mm}
\begin{minipage}{6cm}
  \includegraphics[width=\linewidth]{ActaXFIG/P-QR}
  \caption{Composed pNet for ``P>>(Q>>R)''}  \label{schema:enable-composed}
\end{minipage}

@@

	To give simple intuitions of the open pNet model and its semantics, we use
	here a simple example based on the ``Enable'' operator of the LOTOS language.
	In Fig. \ref{schema:enable-pnets}, we show 2 possible encodings of the  
	``Enable'' operator.
	In the Enable expression ``P>>Q'', an \texttt{exit} statement within P
	terminates the current process, carrying a value $x$ that
	is captured by the \texttt{accept} statement of Q.

	We use a simple action algebra,
	containing two constructors $\delta(x)$ and $acc(x)$, for any
	possible data type of the variable $x$, corresponding to the
	statements \texttt{exit} and \texttt{accept}. Both $\delta(x)$
	and $acc(x)$ actions are
	implicitly included in the sorts of all processes.
	We need no specific predicate over the action expressions, apart
	from equality of actions. 
	In the first encoding $EnableState$, in the upper part of
	Fig. \ref{schema:enable-pnets}, we use a controller $CState$ with two
	states, and simple control actions $l, r, \delta$. The second encoding
	$EnableData$ uses a data-oriented style, with a single state controller $CData$, and a
	state-variable $v$, with values in $\{0,1\}$.

@@

	In Fig. \ref{schema:enable-composed}, we enrich our example by
	composing 2 occurences of the EnableState pNet. In section
        \ref{section:use-case:Enable} we show how these encodings can
        be use to prove properties of the Enable operator.

@@

\begin{example}[Open transitions]
  Transition $OT_1$ of the pNet $EnableStateCompLeft$ is generated by
  combination of the first vector of its root node, with
  the second vector of its $2^{nd}$ node. Its generated predicate is
  trivialy true. The global states shows that controler $CState1$ has
    changed from state 0 to 1.

  \label{OT:example-1}
        \begin{mathpar}
	  OT_1  =
          \inferrule*[myfraction=\reddottedrule]
	{\{\xrightarrow{\delta(x2)}_P ~~
          \xrightarrow{acc(x2)}_Q \} ~~
          [\forall z. acc(x2) \neq \delta(y1)]
        }
	{\ostate{0,0} \xrightarrow {\underline{\delta(x2)}} \ostate{1,0}}
	\end{mathpar}

If we use the data-oriented pNet $EnableData$ from Figure
\ref{schema:enable-pnets} to building a similar $EnableDataCompLeft$
``P>>(Q>>R)'' expression, the corresponding open transition would be:

  \label{OT:example-2}
        \begin{mathpar}
	  OT_2  =
          \inferrule*[myfraction=\reddottedrule]
	{\{\xrightarrow{\delta(x2)}_P ~~
          \xrightarrow{acc(x2)}_Q \} ~~
          [v_1=0 \land v_2 = 0 \land \forall z. acc(x2) \neq
            \delta(z)] ~~ \{v1:=1\}
        }
	{\ostate{0,0} \xrightarrow {\underline{\delta(x2)}} \ostate{0,0}}
	\end{mathpar}
        where $v_1$ and $v_2$ are respectively the state variables of
        the first and second EnableData pNets. Indeed the global
        state space of this version is reduced to $\ostate{0,0}$, and
        the state change is replaced by the assignment ``$v_1:=1$''.

@@

\begin{figure}[t]
        \begin{mathpar}
	  OT_{19}  =
          \inferrule*[myfraction=\reddottedrule]
	{\{\xrightarrow{\delta(x2)}_P ~~
          \xrightarrow{acc(x2)}_R \} ~~
          [\forall z. \delta(x2) \neq \delta(z) \neq
            \delta(x2)]
        }
	{\ostate{2,0} \xrightarrow {\underline{\delta(x2)}} \ostate{3,0}}
	\end{mathpar}
\caption{One of the unsatisfiable open transitions}  \label{schema:unsat-ot}
\end{figure}

Our matching/filtering strategy builds some transitions where
the predicates express incompatible constraints. Even if having an
unsatisfiable (symbolic) transition would not be incorrect,
we choose to minimize the open automaton (\ie its number of
transitions and states), by checking the predicates for satisfiability.
In Fig.~\ref{schema:unsat-ot}, we show an 
unsatisfiable open transition 
from the open automaton of our running example. \TODO{change the text}It shows the case where
the failure controller performs a ``$\PortFail$'' action, while the timer
executes a ``$\PortResume$''. The chosen synchronization
vector (SV0 from Fig.~\ref{schema:ArchFailure:pNet}) does not match
with these actions, since it expects $\NameTimer.\PortStart$. This
  mismatch is materialised by the predicate
fragment ``$\mathit{resume}(\false)=\mathit{start}(b_2)$''.

@@

\begin{algorithm}[h]
  \caption{``Smart'' Reachable Open Automaton Generation}
  \label{alg3}
\begin{algorithmic}[1]
\Require A pNet $P$ (cannot be a hole)
        %\qquad \hl{\textbackslash\textbackslash~SB: was `` pn ''}
\State Initialize sets $U=\{\mathit{InitState}(P)\}$ and $E=\emptyset$,
for unexplored and explored global states, respectively; $L=\emptyset$ for the resulting OTs;
\While{!$\mathit{isEmpty}(U)$}
	\State Choose $S$ in $U$; remove $S$ from $U$, add $S$ to $E$;
        %\qquad \hl{\textbackslash\textbackslash~SB: was `` Choose $S$ in $U$ ''}
	\State $\mathit{OTs} = \mathit{MakeTransitions}(P, S)$;
%%        \State Store $\mathit{OT}$ in $L$; 
	\For {each $\mathit{OT} \in \mathit{OTs}$}
        Check satisfiability of $\mathit{OT}$ using the SMT solver;
        \IfThen {$SAT(\mathit{OT})$}
                {\State \{Add $\mathit{OT}$ to $L$;
                  \State Let $S'$ be the target global state of $\mathit{OT}$
	          \IfThen{($S' \not\in U \cup E$)}{Add $S'$ into $U$;}
                  \}}
	\EndFor
\EndWhile
%%\State Prune the unsatisfiable open transitions in $L$ using the SMT solver;
\IfThenElse {(NonEmptyAssignmentSet($\mathit{L}$))}
        {\State \quad\ \Return $\mathit{OA}$=makeAutomaton($\mathit{InitState}(P),L)$;}
        {\State \quad\ $\mathit{L1}$=filterUnsatTransitions($\mathit{L,Assignments}$);
         \State \quad\ \Return $\mathit{OA}$=makeReachableSubAutomaton($\mathit{InitState}(P),L1)$};
\end{algorithmic}  
\end{algorithm}

@@

Figure~\ref{schema:smt-lib} shows the translation of the
transition of Fig.~\ref{schema:unsat-ot}
in the SMTlib syntax.
\TODO{Change the text}It contains the declaration
of the BIP action algebra sorts 
and constructors, then the declaration of variables, and finally the
predicate to be checked, encoded as a set of assertions.
Here the behavior of the hole ``B'' appears as a variable
``|hb\_B:13:1|'', and line 10 shows the constraint that matching with
a synchronisation vector has imposed on the behavior of ``B''.
Here we display also the diagnosis (``sat'' or ``unsat'') generated by Z3.

\begin{figure}[t]
  %  \centerline{\includegraphics[width=12cm]{XFIG/SMTLIB2}}
    \centerline{\includegraphics[width=0.8\linewidth]{ActaXFIG/rise4fun1}}
  \caption{Checking one Open Transition in Z3}  \label{schema:smt-lib}
\end{figure}

@@

\section{Use-cases}
\label{section:use-cases}
\TODO{new section here, showing various features on different examples}

\subsection{Enable operator} in Lotos, with (equivalent) state-based
and data-based encodings; Associativity of Enable;
\label{section:use-case:Enable}

%%\subsection{simple communication protocol} specification and implementation;

\subsection{BIP FailureTimer architecture}
\label{section:FailureTimerMax}

@@

\subsection{Composing 2 FailureTimer BIP Architectures}
\noteInEM{Modifications stopped here}
\label{section:use-case:Double}
\TODO{Enhanced FailureTimer, 2 instances composed, with duplicated control and Timer}
